# Indian Premiere League Analysis

```{r}
library(tidyverse)
library(esquisse)
library(ggplot2)
library(scales)
library(gt)
library(gtExtras)
library(tidytext)
library(wesanderson)
library(dplyr)
library(tibble)
library(elo)
library(forstringr)
library(reactable)
```

# Data Preparation

```{r}
matches <- read.csv("Matches.csv")
ball_by_ball <- read.csv("Ball-by-Ball.csv")

#Joining both the dataframes using id
ipl = matches %>% inner_join(ball_by_ball, by ="id")

#Converting batsman_runs from string to integer
ipl$batsman_runs = as.numeric(ipl$batsman_runs)

#Creating a variable season by extracting year from date
ipl = ipl %>% mutate(season = str_left(date, 4))
ipl = ipl %>% mutate(season = strtoi(season))

#Renaming teams "Delhi Daredevils" to "Delhi Capitals" & "Deccan Chargers" to "Sunrisers Hyderabad" since only the team names have changed over the seasons to avoid miscalculations.

ipl['bowling_team'][ipl['bowling_team'] == 'Delhi Daredevils'] <- 'Delhi Capitals'
ipl['bowling_team'][ipl['bowling_team'] == 'Deccan Chargers'] <- 'Sunrisers Hyderabad'
ipl['batting_team'][ipl['batting_team'] == 'Delhi Daredevils'] <- 'Delhi Capitals'
ipl['batting_team'][ipl['batting_team'] == 'Deccan Chargers'] <- 'Sunrisers Hyderabad'

matches['team1'][matches['team1'] == 'Delhi Daredevils'] <- 'Delhi Capitals'
matches['team1'][matches['team1'] == 'Deccan Chargers'] <- 'Sunrisers Hyderabad'
matches['team2'][matches['team2'] == 'Delhi Daredevils'] <- 'Delhi Capitals'
matches['team2'][matches['team2'] == 'Deccan Chargers'] <- 'Sunrisers Hyderabad'
matches['toss_winner'][matches['toss_winner'] == 'Delhi Daredevils'] <- 'Delhi Capitals'
matches['toss_winner'][matches['toss_winner'] == 'Deccan Chargers'] <- 'Sunrisers Hyderabad'
matches['winner'][matches['winner'] == 'Delhi Daredevils'] <- 'Delhi Capitals'
matches['winner'][matches['winner'] == 'Deccan Chargers'] <- 'Sunrisers Hyderabad'

#Filtering out matches & teams who played less than 2 seasons
matches <- filter(matches, team1 != "Gujarat Lions")
matches <- filter(matches, team1 != "Rising Pune Supergiant")
matches <- filter(matches, team1 != "Pune Warriors")
matches <- filter(matches, team1 != "Rising Pune Supergiants")
matches <- filter(matches, team1 != "Kochi Tuskers Kerala")
matches <- filter(matches, team1 != "NA")
matches <- filter(matches, team2 != "Gujarat Lions")
matches <- filter(matches, team2 != "Rising Pune Supergiant")
matches <- filter(matches, team2 != "Pune Warriors")
matches <- filter(matches, team2 != "Rising Pune Supergiants")
matches <- filter(matches, team2 != "Kochi Tuskers Kerala")
matches <- filter(matches, team2 != "NA")
matches <- filter(matches, winner != "NA")
```

# Identifying Top 3 Batsmen in Each Team for Season 2020

```{r}
topbatsmen = ipl %>% group_by(batsman,batting_team,season) %>% summarise(batsmanRuns = sum(batsman_runs))
topbatsmen <- topbatsmen [order(-topbatsmen$season, topbatsmen$batting_team, -topbatsmen$batsmanRuns),]
topbatsmen <- filter(topbatsmen, season == "2020")
topbatsmen1 <- topbatsmen %>%  group_by(batting_team) %>%  top_n(3, batsmanRuns)
```

# Plot for Top 3 Batsmen in Each Team by Runs for Season 2020

```{r}
ggplot(topbatsmen1) + aes(x = batsman, y = batsmanRuns, fill = batting_team) + geom_col() + scale_fill_hue(direction = 1) +
 labs(x = "Batsman", y = "Total Runs", title = "Top 3 Batsmen in Each Team by Runs", subtitle = "Season 2020", fill = "Team") +
 coord_flip() + theme_light() + theme(legend.position = "top")
```

# Top Runs Scorers Plot

```{r}
plot1 <- topbatsmen1[order(-topbatsmen1$batsmanRuns),]
plot1 <- filter(plot1, batsmanRuns > "500")
ggplot(plot1) + aes(x = batsman, y = batsmanRuns, fill = batting_team) +
 geom_col() + geom_label(label= plot1$batsmanRuns)+
 scale_fill_viridis_d(option = "cividis", direction = 1) +
 labs(x = "Batsman", y = "Total Runs scored", title = "Top Runs Scorers") +
 theme_minimal()
```

# Identifying Match-ups for Top Batsmen

```{r}
batmatchups = ipl %>% group_by(batsman,bowler) %>% summarise(batsmanRuns = sum(batsman_runs))
batmatchups <- batmatchups [order(-batmatchups$batsmanRuns),]

#Joining topbatsmen1 dataframe with matchups dataframe using batsmen
bats_matchup = topbatsmen1 %>% inner_join(batmatchups, by ="batsman")
bats_matchup = bats_matchup %>% select(batsman, batting_team, bowler, batsmanRuns.y)
bats_matchup <- bats_matchup [order(-bats_matchup$batsmanRuns.y),]
bats_matchup <- bats_matchup %>%  group_by(batsman) %>%  top_n(3, batsmanRuns.y)
```

```{r}
bats_matchup <- bats_matchup [order(bats_matchup$batsman),]

bats_matchup %>% select(c(batsman, batting_team, bowler, batsmanRuns.y)) %>% reactable(defaultColDef = colDef(align = "center"), columns = list(batsman = colDef(name="Batsman"), batting_team = colDef(name="Batting Team"), bowler = colDef(name="Bowler"), batsmanRuns.y = colDef(name="Runs")))
```

# Identifying Top 3 Bowlers in Each Team

```{r}
topbowler <- filter(ipl, dismissal_kind != "NA")
topbowler <- topbowler %>% group_by(bowler) %>%   summarise(wickets=n(), .groups = 'drop')

ipl2 <- ipl [order(ipl$bowler, -ipl$season),] 
ipl2 <- as.data.frame(ipl2)
ipl2 <- ipl2 %>% distinct(bowler, .keep_all = TRUE)
topbowler1 = topbowler %>% inner_join(ipl2, by ="bowler")
topbowler1 = topbowler1 %>% select(bowler, bowling_team, wickets)
topbowler1 <- topbowler1 [order(-topbowler1$wickets),]
topbowler1 <- topbowler1 %>%  group_by(bowling_team) %>%  top_n(3, wickets)

#Removing teams played less than 2 seasons
topbowler1 <- filter(topbowler1, bowling_team != "Gujarat Lions")
topbowler1 <- filter(topbowler1, bowling_team != "Rising Pune Supergiant")
topbowler1 <- filter(topbowler1, bowling_team != "Pune Warriors")
topbowler1 <- filter(topbowler1, bowling_team != "Rising Pune Supergiants")
topbowler1 <- filter(topbowler1, bowling_team != "Kochi Tuskers Kerala")
topbowler1 <- filter(topbowler1, bowling_team != "NA")
```

# Identifying Match-ups for Bowlers

```{r}
matchups_bowler = ipl %>% group_by(bowler,batsman) %>% summarise(wickets = sum(is_wicket))
matchups_bowler = filter(matchups_bowler, wickets != "0")
matchups_bowler <- matchups_bowler [order(-matchups_bowler$wickets),]

#Joining topbowler1 dataframe with matchups_bowler dataframe using bowler
bowler_matchup = topbowler1 %>% inner_join(matchups_bowler, by ="bowler")
bowler_matchup = bowler_matchup %>% select(bowler, bowling_team, batsman, wickets.y)
bowler_matchup1 = bowler_matchup %>% inner_join(topbatsmen1, by ="batsman")
bowler_matchup1 <- bowler_matchup1 [order(-bowler_matchup1$wickets.y, bowler_matchup1$bowler),]
bowler_matchup1 <- bowler_matchup1 %>%  group_by(bowler) %>%  top_n(3, wickets.y)
```

```{r}
bowler_matchup1 <- bowler_matchup1 [order(bowler_matchup1$bowler),]

bowler_matchup1 %>% select(c(bowler, bowling_team, batsman, wickets.y)) %>% reactable(defaultColDef = colDef(align = "center"), columns = list(bowler = colDef(name="Bowler"), bowling_team = colDef(name="Bowling Team"), batsman = colDef(name="Batsman"), wickets.y = colDef(name="Total Dismissals")))

```

# Top Wicket Takers

```{r}
plot2 <- topbowler1[order(-topbowler1$wickets),]
plot2 <- filter(plot2, wickets > 150)

ggplot(plot2) + aes(x = bowler, y = wickets, fill = bowling_team) +
 geom_col() + scale_fill_brewer(palette = "Purples", direction = 1) +
 coord_flip() + geom_text(aes(label = wickets), hjust = 0.25)+
 labs(x = "Bowler", y = "Total Wickets", title = "Top Wicket Takers") +
 theme_minimal()
```


# Win Probability using ELO

```{r}
# Creating a new data frame to hold only Unique teams
teams = data.frame(team = unique(c(matches$team1, matches$team2)))

# Number of matches played by each team
played1 <- matches %>%  group_by(team1) %>%  summarize(count1 = n()) %>%
  arrange(team1) %>%  rename(team = team1)

played2 <- matches %>%  group_by(team2) %>%  summarize(count2 = n()) %>%
  arrange(team2) %>%  rename(team = team2)

matches_team <- played1 %>%  full_join(played2, by = "team") %>%
  mutate(n_matches_played = count1 + count2) %>%
  select(team, n_matches_played) %>% arrange(desc(n_matches_played))
matches_team

matches_team %>%  ggplot(aes(team, n_matches_played, fill = team)) +
  geom_col(show.legend = FALSE) +  coord_flip() +
  geom_text(aes(label = n_matches_played), hjust = 0.25)+
  labs(title="Number of Matches Played by Each Team", x= "Team Names", y = "Number of Matches Played")
```
# Identifying number of matches won by each team

```{r}
plot <- matches %>% select(winner) 
plot <- plot %>% group_by(winner) %>% summarise(matches_won=n(),
            .groups = 'drop')
plot <- plot[order(-plot$matches_won),]

plot %>%  ggplot(aes(winner, matches_won, fill = winner)) +
  geom_col(show.legend = FALSE) +  coord_flip() +
  geom_text(aes(label = matches_won), hjust = 0.25)+
  labs(title="Number of Matches Won by Each Team", x= "Team Names", y = "Number of Matches Won")
```


```{r}
#Assigning initial ratings to all teams as 1,500
teams = teams %>% mutate(elo = 1500)
matches = matches %>% mutate(resultPoint = if_else(team1 == winner, 1, if_else(team2 == winner, 0, 0.5)))
matches <- matches %>% drop_na(resultPoint)
matches1 = matches %>% select(date, team1, team2, winner, resultPoint) %>% arrange(date)

for (i in seq_len(nrow(matches1))) {
  match = matches1[i, ]
  
  # Pre-match ratings
  teamA_elo = subset(teams, team == match$team1)$elo 
  teamB_elo = subset(teams, team == match$team2)$elo
  # Let's update our ratings
  new_elo = elo.calc(wins.A = match$resultPoint, 
                     elo.A = teamA_elo, 
                     elo.B = teamB_elo,
                     k = 30)
  
  # The results come back as a data.frame
  # with team A's new rating in row 1 / column 1 
  # and team B's new rating in row 1 / column 2 
  teamA_new_elo = new_elo[1, 1]
  teamB_new_elo = new_elo[1, 2]
  
  # We then update the ratings for teams A and B 
  # and leave the other teams as they were
  teams = teams %>%
    mutate(elo = if_else(team == match$team1, teamA_new_elo,
                         if_else(team == match$team2, teamB_new_elo, elo)))
}
teams %>%  arrange(-elo) %>%  head

teams %>% ggplot(aes(team, elo, fill = team)) + geom_col(show.legend = FALSE) +
  coord_flip() + geom_text(aes(label = elo), hjust = 0.25)+
  labs(title="ELO Score of Each Team after 13 Seasons", x= "Team Names", y = "ELO Score")
```

```{r}
Mumbai_Indians <- subset(teams, team == "Mumbai Indians")$elo 
Chennai_Super_Kings <- subset(teams, team == "Chennai Super Kings")$elo
Kolkata_Knight_Riders <- subset(teams, team == "Kolkata Knight Riders")$elo 
Delhi_Capitals <- subset(teams, team == "Delhi Capitals")$elo
Sunrisers_Hyderabad <- subset(teams, team == "Sunrisers Hyderabad")$elo 
Rajasthan_Royals <- subset(teams, team == "Rajasthan Royals")$elo
Kings_XI_Punjab <- subset(teams, team == "Kings XI Punjab")$elo 
Royal_Challengers_Bangalore <- subset(teams, team == "Royal Challengers Bangalore")$elo

runs = 1000 #number of simulations to run

game = function(team1,team2,team1name,team2name){
  prob = elo.prob(team1, team2)
  rand = runif(1)
  winner = ifelse(prob > rand, team1name, team2name)
}

set.seed(123)
csk_srh = replicate(runs,game(Chennai_Super_Kings,Sunrisers_Hyderabad,"Chennai Super Kings","Sunrisers Hyderabad"),simplify=FALSE) %>%
  do.call(rbind.data.frame,.)

set.seed(123)
csk_rcb = replicate(runs,game(Chennai_Super_Kings,Royal_Challengers_Bangalore,"Chennai Super Kings","Royal Challengers Bangalore"),simplify=FALSE) %>%  do.call(rbind.data.frame,.)

set.seed(123)
csk_rr = replicate(runs,game(Chennai_Super_Kings,Rajasthan_Royals,"Chennai Super Kings","Rajasthan Royals"),simplify=FALSE) %>%
  do.call(rbind.data.frame,.)

set.seed(123)
csk_mi = replicate(runs,game(Chennai_Super_Kings,Mumbai_Indians,"Chennai Super Kings","Mumbai Indians"),simplify=FALSE) %>%
  do.call(rbind.data.frame,.)

set.seed(123)
csk_kkr = replicate(runs,game(Chennai_Super_Kings,Kolkata_Knight_Riders,"Chennai Super Kings","Kolkata Knight Riders"),simplify=FALSE) %>%
  do.call(rbind.data.frame,.)

set.seed(123)
csk_kxp = replicate(runs,game(Chennai_Super_Kings,Kings_XI_Punjab,"Chennai Super Kings","Kings XI Punjab"),simplify=FALSE) %>%
  do.call(rbind.data.frame,.)

set.seed(123)
csk_dc = replicate(runs,game(Chennai_Super_Kings,Delhi_Capitals,"Chennai Super Kings","Delhi Capitals"),simplify=FALSE) %>%
  do.call(rbind.data.frame,.)

csk_league = bind_cols(csk_srh,csk_rcb,csk_rr,csk_mi,csk_kkr,csk_kxp,csk_dc)

head(csk_srh)
head(csk_rcb)
head(csk_rr)
head(csk_mi)
head(csk_kkr)
head(csk_kxp)
head(csk_dc)

csk_league = csk_league %>%
  rename(csk_srh =  c..Chennai.Super.Kings....Sunrisers.Hyderabad....Chennai.Super.Kings...) %>%
  rename(csk_rcb = c..Chennai.Super.Kings....Royal.Challengers.Bangalore....Chennai.Super.Kings...) %>%
  rename(csk_rr = c..Chennai.Super.Kings....Rajasthan.Royals....Chennai.Super.Kings...) %>%
  rename(csk_mi = c..Chennai.Super.Kings....Mumbai.Indians....Mumbai.Indians...) %>%
  rename(csk_kkr = c..Chennai.Super.Kings....Kolkata.Knight.Riders....Chennai.Super.Kings...) %>%
  rename(csk_kxp = c..Chennai.Super.Kings....Kings.XI.Punjab....Chennai.Super.Kings...) %>%
  rename(csk_dc = c..Chennai.Super.Kings....Delhi.Capitals....Chennai.Super.Kings...)
  
head(csk_league)

table(csk_league$csk_srh)
table(csk_league$csk_rcb)
table(csk_league$csk_rr)
table(csk_league$csk_mi)
table(csk_league$csk_kkr)
table(csk_league$csk_kxp)
table(csk_league$csk_dc)
```

# Impact Players

```{r}
IP <- ipl %>% select(id, team1, team2, batsman, batsman_runs, batting_team, bowler, dismissal_kind, bowling_team,season)

#Assigning Impact Points for Batsman for the runs scored. Each Runs will get 1 point with double point option for 6 runs.
IP <- IP %>%  mutate(batting_points = case_when(batsman_runs == 0 ~ 0, batsman_runs == 1 ~ 1,
  batsman_runs == 2 ~ 2,  batsman_runs == 4 ~ 4, batsman_runs == 6 ~ 12))

#Assigning Impact Points for Bowlers for wickets taken. Each wicket is worth 30 points expect for Runs outs & Retired Hurt as it is not applicable to an individual bowler.
IP <- IP %>% mutate(bowling_points = case_when(dismissal_kind == "NA" ~ 0, dismissal_kind == "caught" ~ 30, dismissal_kind == "lbw" ~ 30, dismissal_kind == "caught and bowled" ~ 30, dismissal_kind == "bowled" ~ 30, dismissal_kind == "stumped" ~ 30, dismissal_kind == "run out" ~ 0, dismissal_kind == "retired hurt" ~ 0))

```


#Impact Players - Batsman

```{r}
# For batting:
IP_batting <- IP %>% select(id, team1, team2, batsman, batting_points, batting_team, season)
IP_batting <- filter(IP_batting, season >= "2018")
IP_batting <- IP_batting %>% group_by(id, season, team1, team2, batting_team, batsman) %>% summarise(batting_points = sum(batting_points))
IP_batting1 <- IP_batting %>%  group_by(batting_team, id) %>%  top_n(1, batting_points)
IP_batting1 <- IP_batting1 %>% group_by(season, batting_team) %>% filter(batting_points == max(batting_points, na.rm = TRUE))

IP_batting1 %>% select(c(id, season, team1, team2, batsman, batting_team, batting_points)) %>% reactable(defaultColDef = colDef(align = "center"), columns = list(id = colDef(name="ID"), season = colDef(name="Season"), team1 = colDef(name="Home Team"), team2 = colDef(name="Away Team"), batsman = colDef(name="Batsman"), batting_team = colDef(name="Batting Team"), batting_points = colDef(name="Points")))
```

#Impact Players - Bowler
```{r}
IP_bowling <- IP %>% select(id, team1, team2, bowler,bowling_points, bowling_team, season)
IP_bowling <- filter(IP_bowling, bowling_points != "NA")
IP_bowling <- filter(IP_bowling, season >= "2018")
IP_bowling <- IP_bowling %>% group_by(id, season, team1, team2, bowling_team, bowler) %>% summarise(bowling_points = sum(bowling_points))
IP_bowling1 <- IP_bowling %>%  group_by(bowling_team, id) %>%  top_n(1, bowling_points)
IP_bowling1 <- IP_bowling1 %>% group_by(season, bowling_team) %>% filter(bowling_points == max(bowling_points, na.rm = TRUE))
IP_bowling1 %>% select(c(id, season, team1, team2, bowler, bowling_team, bowling_points)) %>% reactable(defaultColDef = colDef(align = "center"), columns = list(id = colDef(name="ID"), season = colDef(name="Season"), team1 = colDef(name="Home Team"), team2 = colDef(name="Away Team"), bowler = colDef(name="Bowler"), bowling_team = colDef(name="Bowling Team"), bowling_points = colDef(name="Points")))
```